# Data transformation


```{r eval=FALSE, echo=TRUE}
# this prevents package loading message from appearing in the rendered version of your problem set
knitr::opts_chunk$set(echo = TRUE, warning = FALSE,
                      message = FALSE)
library(tidyverse)
library(ggplot2)
library(dplyr)
library(lubridate)
library(scales)
library(choroplethr)
library(choroplethrMaps)
library(Hmisc)
library(ggthemes)
library(viridis)
library(geofacet)
```


set working directory where .csv files are saved
```{r eval=FALSE, echo=TRUE}
# rm(list=ls())
setwd('C:/Users/Administrator.WFC-TRD-WS20/Desktop/DSI - Spring 21/STAT GR5702 EDAV/Final Project')
```


data download
```{r eval=FALSE, echo=TRUE}
covid <- read.csv("covid_confirmed_usafacts.csv", check.names=FALSE)
county_population <- read.csv('covid_county_population_usafacts.csv')
```


covid data at county level (cumulative cases)
```{r eval=FALSE, echo=TRUE}
covid.county <- covid[covid$countyFIPS!=0,]
covid.county  <- pivot_longer(covid.county,
                              cols=5:443,names_to="Date",values_to="Positive") 
covid.county$Date <- as.Date(covid.county$Date, format="%Y-%m-%d")
colnames(covid.county)[1] <- "County.FIPS"
colnames(covid.county)[2] <- "County.Name"

colnames(county_population)[1] <- "County.FIPS"
county_population <- county_population[county_population$County.FIPS!=0,]

covid.county.merged <- merge(covid.county, county_population[c("County.FIPS", "population")], by="County.FIPS")
```


covid data at county level (cumulative cases), cases converted to percentage. county is sub-grouped into 3 sizes
```{r eval=FALSE, echo=TRUE}
describe(county_population$population)

b <- c(-Inf, 10890, 213548, Inf)
names <- c('Small', 'Medium', 'Large')
covid.county.merged$population.cat <- cut(covid.county.merged$population, breaks=b, labels=names)

covid_county_pct <- covid.county.merged %>% 
  mutate(Positive = Positive / population * 100) %>% 
  mutate(across(where(is.numeric), ~ round(., 2)));
```

covid data at county level, now positive case # shows net daily change
```{r eval=FALSE, echo=TRUE}
covid_county_net <- covid.county.merged %>% 
  group_by(County.FIPS) %>% 
  mutate(lag=lag(Positive)) %>% 
  mutate(Positive.change = Positive - lag)
covid_county_net <- covid_county_net[covid_county_net$Date!="2020-01-22",]
```


covid data at state level (cumulative cases), cases converted to percentage. state is sub-grouped into 3 sizes
```{r eval=FALSE, echo=TRUE}
covid.state <- covid.county.merged %>% 
  group_by(Date,State) %>% 
  dplyr::summarise(Positive_count = sum(Positive))

covid.population <- covid.county.merged %>% 
  group_by(Date,State) %>% 
  dplyr::summarise(State_population = sum(population))

covid.merged <- merge(covid.state, covid.population)

describe(covid.merged$State_population)
b <- c(-Inf, 1787065, 12671821, Inf)
names <- c('Small', 'Medium', 'Large')
covid.merged$population.cat <- cut(covid.merged$State_population, breaks=b, labels=names)

covid_state_pct <- covid.merged %>% 
  mutate(Positive_count = Positive_count / State_population * 100) %>% 
  mutate(across(where(is.numeric), ~ round(., 2)));

colnames(covid_state_pct)[3] <- "Positive_pct"
```


covid data at state level, now positive case # shows net daily change
```{r eval=FALSE, echo=TRUE}
covid_state_net <- covid.merged %>% 
  group_by(State) %>% 
  mutate(lag=lag(Positive_count)) %>% 
  mutate(Positive.change = Positive_count - lag) 
covid_state_net <- covid_state_net[covid_state_net$Date!="2020-01-22",]
```


covid data monthly at both county / state level
```{r eval=FALSE, echo=TRUE}
covid_state_monthly <- covid_state_net %>% 
  group_by(State, population.cat, month=lubridate::floor_date(Date, "month")) %>% 
  dplyr::summarise(positive_monthly=sum(Positive_count))

covid_state_pct_monthly <- covid_state_pct %>% 
  group_by(State, population.cat, month=lubridate::floor_date(Date, "month")) %>% 
  dplyr::summarise(positive_pct_monthly=max(Positive_pct))

covid_county_monthly <- covid_county_net %>% 
  group_by(County.Name, State, population.cat, month=lubridate::floor_date(Date, "month")) %>% 
  dplyr::summarise(positive_monthly=sum(Positive.change))

covid_county_monthly <- covid_county_net %>% 
  group_by(County.Name, State, population.cat, month=lubridate::floor_date(Date, "month")) %>% 
  dplyr::summarise(positive_monthly=sum(Positive.change))

covid_county_pct_monthly <- covid_county_pct %>% 
  group_by(County.Name, State, population.cat, month=lubridate::floor_date(Date, "month")) %>% 
  dplyr::summarise(positive_pct_monthly=max(Positive))

covid_state_cum_monthly <- covid.merged %>% 
  group_by(State, population.cat, month=lubridate::floor_date(Date, "month")) %>% 
  dplyr::summarise(positive_monthly=max(Positive_count))
```


data save
```{r eval=FALSE, echo=TRUE}
save(covid_county_pct, 
     covid_county_net, 
     covid_state_pct, 
     covid_state_net, 
     covid_state_monthly,
     covid_state_pct_monthly,
     covid_county_monthly,
     covid_county_pct_monthly,
     covid_state_cum_monthly,
     file = 'covid_cases_v5.Rda')
```


```{r eval=FALSE, echo=TRUE}
library(tidyr)
```


Get the Trips by Distance 

```{r eval=FALSE, echo=TRUE}
T_by_D <- read.csv("Trips_by_Distance.csv")

T_by_D$Date <- as.Date(T_by_D$Date, format="%Y/%m/%d")

trips <- T_by_D[T_by_D$Date >= "2020-01-22" ,]
trips <- subset(trips, select=-Row.ID)
colnames(trips)[c(4,6)] <- c("State","County")

#remove NA rows
#trips <- trips[!(is.na(trips$Number.of.Trips)),]
```

Seperate the trips by National, State, and County levels

```{r eval=FALSE,echo=TRUE}
trips.national <- subset(trips,Level=="National", select= -c(State.FIPS,State,
                                                              County.FIPS,County))
trips.state <- subset(trips,Level=="State", -c(Level,County.FIPS,County))
trips.county <- subset(trips,Level=="County", -Level)
```

save the trip data 

```{r eval=FALSE,echo=TRUE}
save(trips.national,trips.state,trips.county, file="trips.RData")
```


Get the Covid Rates data

```{r eval=FALSE,echo=TRUE}
covid <- read.csv("covid_confirmed.csv", check.names=FALSE)

covid.county <- covid[covid$countyFIPS!=0,]

covid.county  <- pivot_longer(covid.county,
                             cols=5:443,names_to="Date",values_to="Positive") 
covid.county$Date <- as.Date(covid.county$Date, format="%Y-%m-%d")
colnames(covid.county)[c(1:2,4)] <- c("County.FIPS", "County","State.FIPS")
```

Split covid data into State and county sets 

```{r eval=FALSE,echo=TRUE}
covid.state <- covid.county %>% 
  group_by(State, State.FIPS, Date) %>% 
  summarise(Positive = sum(Positive)) %>% 
  mutate(New = ifelse(Positive-lag(Positive)>0,Positive-lag(Positive),0)) %>% 
  replace(is.na(.),0)

covid.county <- covid.county %>% 
  group_by(State, State.FIPS, County, County.FIPS, Date) %>% 
  summarise(Positive = sum(Positive)) %>% 
  mutate(New = ifelse(Positive-lag(Positive)>0,Positive-lag(Positive),0)) %>% 
  replace(is.na(.),0) 
```

save it 

```{r eval=FALSE, echo=TRUE}
save(covid.state,covid.county, file="covid.RData")
```